Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-01 LaTeX/Turnover.tex
discussion, getting closer...

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \textcolor{red}{(Ubeda and Haig, 2005; Lindholm et al. 2016)}. 
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Haploid competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before haploid competition (see Sup.\ Mat. for recursion equations). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during haploid competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during haploid competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Haploid competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
Random mating then occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific haploid competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during haploid competition \\ [0.5ex] \hline
  A & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  a & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  Aa & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  aa & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Tranmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  a & $(1-\alpha^\Hermaphrodite)=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $D^\Hermaphrodite=\big{(} \bar{p}s^\Hermaphrodite+(1-\bar{p})h^\Hermaphrodite s^\Hermaphrodite\big{)} -\big{(} \bar{p}h^\Hermaphrodite s^\Hermaphrodite+(1-\bar{p}) \big{)}$ corresponds to the difference in fitness between $A$ and $a$ alleles in diploids of sex $\Hermaphrodite \in \{\female,\male\}$ ($\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex). 
The frequency of Y among male gametes depends upon the difference in $A$ allele frequency on X- and Y-bearing male gametes and the strength of meiotic drive in favour of the $A$ allele in males, $q=1/2+\alpha_{\Delta}^\male(\hat{p}^\male_Y-\hat{p}^\male_X)/2+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^\Hermaphrodite=t^\Hermaphrodite=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

@@

\begin{threeparttable}[ht]
\centering
\smallskip
\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male}+(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male}\alpha^\male(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male}+p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1-\alpha^\male)(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1-\alpha^{\male})/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes 
      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during haploid competition
      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during haploid competition
       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $S_{A}=(D^\male +\alpha_{\Delta}^\male+t^\male) - (D^\female+\alpha_{\Delta}^\female+t^\female)$ is the difference in fitness in males versus females for the $A$ allele against the $a$ allele across diploid selection, haploid competition, and meiosis.

@@

We find that neo-W alleles can invade for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in \eqref{eq:lambda_neoW} disappears and invasion depends on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region, Figure \ref{fig:Combination_Centimorgans}.
Secondly, we can simplify cases where $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and equal dominance, $h^\female=h^\male$. 
Where there is no haploid competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and haploid competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type ($-s^\female(s^\female-s^\male)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if $R>r$.

@@

%Sex ratios
Previous research suggests, when the ancestral sex-determining locus is linked to a locus that experiences haploid selection (e.g., meiotic drive), a new, unlinked sex-determining locus invades in order to restore equal sex ratios \citep{Kozielska:2010vm}. 
Our model provides a good opportunity to determine whether Fisherian sex ratio selection provides a useful explanation for the evolution of new sex-determining loci in other contexts. 
Consider, for example, the case where the \textbf{A} locus is linked to the ancestral-SDR ($r<1/2$) and experiences meiotic drive in males only (e.g., during spermatogenesis but not during oogenesis, $\alpha^\male \neq 1/2$, $\alpha^\female=1/2$). 
We will also disregard haploid competition ($t^\female=t^\male=0$) such that zygotic sex ratios can only be biased by meiotic drive in males. 
In this case, the zygotic sex ratio can be initially biased only if the ancestral sex-determining system is XY (Figure \ref{fig:Combination_Turnover}B). 
If the ancestral sex-determining system is ZW, the zygotic sex ratio will be 1:1 because diploid sex is determined by the proportion of Z-bearing versus W-bearing eggs (and meiosis in females is fair, Figure \ref{fig:Combination_Turnover}D).
Thus, if the zygotic sex ratio is crucial to the evolution of new genetic sex-determining systems, invasion into ZW and XY systems will be distinct. 
However, we find that invasion by a homogametic neo-sex-determining allele (XY to XY, or ZW to ZW) or by a heterogametic neo-sex-determining allele (XY to ZW or ZY to XY) occur under the same conditions. 
That is, we can show that $\lambda_{Y',XY}=\lambda_{W',ZW}$ and $\lambda_{Y',ZW}=\lambda_{W',XY}$, for a numerical example, compare Figure \ref{fig:Combination_Turnover}A,B to Figure \ref{fig:Combination_Turnover}C,D.

@@

From vD\&K 2010:
``We noted in the Introduction that several recently derived sex chromosomes are known with genes that appear to be targets of sex-antagonistic selection (Kallman 1973; Wada et al. 1998; Lande et al. 2001; Lindholm and Breden 2002; Streelman et al. 2003; Fernandez and Morris 2008; Kitano et al. 2009; Roberts et al. 2009). These observations are certainly consistent with the hypothesis that sex-antagonistic selection was responsible for the establishment of a new sex-determining system, but a proper test of our hypothesis would require data to rule out the plausible alternative that the sex-antagonistic alleles accumulated secondarily, after the new sex chromosomes invaded. Indeed, linkage to sex chromosomes is favorable to the maintenance of sex-antagonistic variation (Rice 1987). The required data may soon become available through comparative genomic studies on species that recently underwent a heterogamety transition and their close relatives. The presence of sex-antagonistic variation at the same genomic location in closely related species that have the ancestral sex-determining system would support the view that sex-antagonistic variation predated the origin of the derived sex chromosomes and caused the invasion of a new sex-determination system.
''

Note that, ploidally-antagonistic selection does not suffer from the proviso that it is easier to maintain polymorphisms close to the SDR.  vD\&K also suggest that this build up of sex-antagonistic polymorphisms may help to stabilize the ancestral sex-determining system, which would not be the case with haploid selection.

@@

Among gametes from sex $\Hermaphrodite$ (sperm/pollen when $\Hermaphrodite=\male$, eggs/ovules when $\Hermaphrodite=\female$), the frequency of haplotypes (before haploid competition) in the next generation are given by

>>>>

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \textcolor{red}{(Ubeda and Haig, 2005; Lindholm et al. 2016)}. 
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition (see Sup.\ Mat. for recursion equations). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  A & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  a & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  Aa & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  aa & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Tranmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  a & $(1-\alpha^\Hermaphrodite)=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $D^\Hermaphrodite=\big{(} \bar{p}s^\Hermaphrodite+(1-\bar{p})h^\Hermaphrodite s^\Hermaphrodite\big{)} -\big{(} \bar{p}h^\Hermaphrodite s^\Hermaphrodite+(1-\bar{p}) \big{)}$ corresponds to the difference in fitness between $A$ and $a$ alleles in diploids of sex $\Hermaphrodite \in \{\female,\male\}$ ($\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex). 
The frequency of Y among male gametes depends upon the difference in $A$ allele frequency on X- and Y-bearing male gametes and the strength of meiotic drive in favour of the $A$ allele in males, $q=1/2+\alpha_{\Delta}^\male(\hat{p}^\male_Y-\hat{p}^\male_X)/2+O(\epsilon^3)$.
Without gametic competition or drive ($\alpha_{\Delta}^\Hermaphrodite=t^\Hermaphrodite=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

@@

\begin{threeparttable}[ht]
\centering
\smallskip
\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male}+(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male}\alpha^\male(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male}+p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1-\alpha^\male)(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1-\alpha^{\male})/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes 
      \item see Table \ref{tab:meanfitnesses} for mean fitnesses.
%      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during gametic competition
%      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during gametic competition
%       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
%       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $S_{A}=(D^\male +\alpha_{\Delta}^\male+t^\male) - (D^\female+\alpha_{\Delta}^\female+t^\female)$ is the difference in fitness in males versus females for the $A$ allele against the $a$ allele across diploid selection, gametic competition, and meiosis.

@@

We find that neo-W alleles can invade for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in \eqref{eq:lambda_neoW} disappears and invasion depends on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region, Figure \ref{fig:Combination_Centimorgans}.
Secondly, we can simplify cases where $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and equal dominance, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type ($-s^\female(s^\female-s^\male)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if $R>r$.

@@

%Sex ratios
Previous research suggests, when the ancestral sex-determining locus is linked to a locus that experiences haploid selection (e.g., meiotic drive), a new, unlinked sex-determining locus invades in order to restore equal sex ratios \citep{Kozielska:2010vm}. 
Our model provides a good opportunity to determine whether Fisherian sex ratio selection provides a useful explanation for the evolution of new sex-determining loci in other contexts. 
Consider, for example, the case where the \textbf{A} locus is linked to the ancestral-SDR ($r<1/2$) and experiences meiotic drive in males only (e.g., during spermatogenesis but not during oogenesis, $\alpha^\male \neq 1/2$, $\alpha^\female=1/2$). 
We will also disregard gametic competition ($t^\female=t^\male=0$) such that zygotic sex ratios can only be biased by meiotic drive in males. 
In this case, the zygotic sex ratio can be initially biased only if the ancestral sex-determining system is XY (Figure \ref{fig:Combination_Turnover}B). 
If the ancestral sex-determining system is ZW, the zygotic sex ratio will be 1:1 because diploid sex is determined by the proportion of Z-bearing versus W-bearing eggs (and meiosis in females is fair, Figure \ref{fig:Combination_Turnover}D).
Thus, if the zygotic sex ratio is crucial to the evolution of new genetic sex-determining systems, invasion into ZW and XY systems will be distinct. 
However, we find that invasion by a homogametic neo-sex-determining allele (XY to XY, or ZW to ZW) or by a heterogametic neo-sex-determining allele (XY to ZW or ZY to XY) occur under the same conditions. 
That is, we can show that $\lambda_{Y',XY}=\lambda_{W',ZW}$ and $\lambda_{Y',ZW}=\lambda_{W',XY}$, for a numerical example, compare Figure \ref{fig:Combination_Turnover}A,B to Figure \ref{fig:Combination_Turnover}C,D.

@@

vD\&K also suggest that this build up of sex-antagonistic polymorphisms may help to stabilize the ancestral sex-determining system, which would not be the case with haploid selection.

@@

Among gametes from sex $\Hermaphrodite$ (sperm/pollen when $\Hermaphrodite=\male$, eggs/ovules when $\Hermaphrodite=\female$), the frequency of haplotypes (before gametic competition) in the next generation are given by
